Do Machines Think Equally About Music?

### 2025-05-05 paper/paper.qmd
update paper based on periklis feedback; now we are done!

Recently, AI-powered tools like Suno [@suno] have emerged, targeting consumers interested in "creating" their own music. This raises questions about how machines understand music, especially in the case of generative models such as Suno's. While more research-based proofs of concept like Meta's MuseGen [@musegen] have existed for years, tools like Suno adopt a more consumer-focused approach. But how does a word entered into a prompt box translate into the auditory representation humans understand as music? Do different artificial intelligence systems have different conceptualizations of commonly accepted human concepts of genre?

To understand if there is a different between conceptualizations of genre between different machine learning systems, we train several different machine learning algorithms in the task of predicting genre, based on the dataset of the timbral audio features of songs across 10 genres extract from the Free Music Archive [@fma]. Then, we ask the 'small' variant of Meta MuseGen to create 'quintessential' representations of different genres of music using systematic prompts, and test if the algorithm can accurately predict the genre based on extracted audio features.

The most balanced classifier, a Support Vector Machine (SVM) model, was unable to guess the genre of any of the machine-generated tracks. It is unclear why is this. It could either be because of the perhaps due to the music generation model's inability to create music that follows patterns of any genre, or it may be because the SVM performs very poorly. In this case, the poor performance of the SVM inhibits deeper insight, but does signal that diffrent machine learning models "learn" the conceptualization of particular music genres and the features that make it up

@@

As @tzanetakis2002, @chen2009, and @bhalke2017 found, timbral metrics allow for sophisticated and parsimonious audio analyses, allowing for efficient classification of genre across various genres of music, and so they are made central to the modelling strategy described in @sec-modelling

@@

After understanding timbral features' centrality in genre classification, finding extensive audio feature data to train the classifier was then made of interest. A key requirement of this data is that all music sourced within it must be human generated, and so we use the Free Music Archive's (FMA's) collection [@fma]. The full archive contains 106,574 tracks across 163 subgenres, and was published in 2017, predating significant public interest in artificial music generation, so it is reasonably unlikely that any model-based music was incorporated into the dataset. While the FMA is technically made of audio tracks, which require large amounts of memory to store, Defferard et al. also extracted audio features from each of the tracks using Python's `librosa` module [@librosa], as well as provided metadata for each of the songs, particularly a unique ID, title, artist names, and primary genre of the music.

The data cleaning process began by loading raw audio feature data. From this, the key timbral features of each track were extractedd and consolidated systematically into a single, new dataset. The feature groups that were extracted were the first 16 mean values for the Mel-Frequency Cepstral Coefficients (MFCCs), and the means and variances for the spectral centroid, spectral rolloff, and zero crossing rates for each track. Spectral Flux was not one of the coefficients recorded by Defferard et al, and so is omitted from the feature set.

@@

The dataset was then visualized to understand the distribution of the outcome variable, genre.

@@

As @fig-genre-bar shows, there exists a large amount of bias in the preliminary dataset toward certain primary genres.  A large amount of the observations tended to have a primary genre of Rock, followed by experimental and electronic music. The rest of the 16 primary genres have far less observations. Given this, the training data was refined by restricting the analysis to only the top 10 most frequently occurring genres. With respect to the table, this corresponds to the genres of:

@@

By reducing the number of classes to 10, the classification task is simplified and more targeted comparisons and analysis can be made between classifying human and machine-generated music. Though it is still certainly unbalanced, as seen in @tbl-genre-counts, class-weighting is utilized in the modelling process to account for the overrepresentation of certain genres.

With respect to the predictor variables of the classifier, while it would be possible to visualize each coefficient of the MFCC spectrum, centroids, and other features, each individual coefficient of a song's MFCCs and spectral features are not particularly useful in gleaning special insight on the dataset as a whole. Thus, visualizing the feature set is of adds little to understanding the general aspects of the auditory structure before modelling it and so it remains unvisualized.

@@

Python's `scikit-learn` [@scikit] and `xgboost` [@xgboost] modules were used to implement several classification models, which were iteratively trained and tested before settling on the one with the best performance. Before training each model, the dataset was randomly split into a train-test split of 80% training to 20% testing. In order of least to best test performance, the Support Vector Machine (SVM), K-nearest neighbours, Random Forest, and Extreme Gradient Boosting were evaluated. Due to restrictions in computing power, Deep Learning models were not able to be tested, but were initially considered for this problem. The Support Vector Machine (SVM) model, while the least accurate overall, was the most balanced in prediction across genres, and so was then used to classify the artificially generated music described in @sec-results. Detailed diagnostics for each model tested are included in @sec-model-diag.

@@

Support Vector Machines (SVMs) are a widely used supervised learning method for classification tasks, such as identifying the genre of a piece of music. SVMs work by finding the best possible boundary, or hyperplane, that separates songs from different genres based on the extracted timbral audio features. When the relationship between genres is too complex for a simple linear boundary, SVMs can apply kernel functions to capture more intricate, non-linear patterns in the data and find higher-dimensional boundaries between the features, effectively classifying them into genres. By focusing mainly on the most informative examples in the dataset, known as support vectors, SVMs achieve strong performance while avoiding overfitting the training data [@islp]. This makes them particularly effective for classifying music, where genre characteristics often overlap and vary in subtle ways.

Both @chen2009 and @bhalke2017 demonstrated particular success in genre classification with SVM, each implementing them to predict genre from audio features to a high degree of accuracy. Bhalke et al. demonstrated nearly 96% accuracy on their music classification task, so it was expected that SVMs would exhibit similar good performance using the FMA data. The setup for our SVM training used a RBF kernel, with class-weighting enabled in order to account for the unbalancedness of the dataset. Results of testing on the test set are presented in @tbl-svm-results

@@

Surprisingly, this classifier had an accuracy of about 49%, which prompted the investigation of other classification methods for this task described later on. More surprisingly, however was how effectively the SVM learned the patterns of Classical music, despite it having the 2nd least amount of training examples in the dataset, achieving recall of nearly 80% and an F1 score of 0.6. Upon further reflection, this makes sense, as many subgenres of 'Classical' music have particular blends of instruments whose sounds combine together in predictable ways. An example of this within the context of the dataset is that of string quartets, whose timbre of the 2 Violins, Viola, and Cello come together in predictable ways across recordings, which the classifier then learns.

On the other hand, the genres of Jazz and Pop perform relatively poorly, even with class-weighting, with F1 scores of less than 20 percent. Ranking the relative feature importances and then retraining them on the top 10 most importance features did not improve the accuracy by much, either. Even so, the SVM trained is a decently balanced classification model.

@@

Next, the K-nearest neighbours algorithm (KNN) was tested. @tzanetakis2002 had some modest success using the KNN classifier, with above-50%-accurate genre prediction based on audio features. KNN classifies a new song based on the genres of its K closest neighbors in the feature space, considering characteristics like tempo, rhythm patterns, and timbre. The number of neighbors, K, is a key parameter that balances flexibility and stability: smaller values of K make the model sensitive to local patterns, while larger values smooth out noise but may overlook finer distinctions. KNN does not build an explicit model but instead relies directly on the training data, making it highly adaptable but potentially sensitive to the choice and scaling of features [@islp]. In music genre classification, KNN is particularly useful when songs of the same genre tend to cluster closely together based on their properties. Given that music of a certain genre tends to "sound" similar in terms of timbre, it would make sense that KNN could work well at classifying genre based on how those features cluster together.

The setup for our KNN classifier was such that based on a $K = 25$, which was found to balance noise control with precision. Testing results are presented in @tbl-knn-results

@@

This classifier had an overall accuracy of about 54%, improving it slightly beyond what the SVM could predict overall. Classical Music, despite being less represented That said, given that classes could not be weighted, the lesser-represented classes suffered greatly in learning quality. So, from the SVM to KNN, we observed correct classification rates for Pop music go from 16.3% (SVM) to 6.8% (KNN) and Jazz from 14.7% to 3.4%, making this classifier much more biased than the SVM. More represented genres tended to perform even better than in the SVM, with correct Electronic music classification going from 51.9% to 55.3%.

@@

Next, tree-based classification methods were then attempted. A Random Forest classifier was attempted first.

@@

The Random Forest was implemented using class-weighting to address the biased data, with the final constructed tree having 22,023 nodes and a depth of 32. Classification results are presented in @tbl-rf-results.

@@

The XGBoost model was also implemented using class-weighting to account for the biased data, with the final constructed ensemble consisting of 100 different decision trees.

@@

This model further improved upon the Random Forest in overall accuracy, with a accuracy of 56.4% overall. XGBoost did marginally improve classification rates of the less represented genres of Jazz and Pop. Correct classification of jazz increased from 8.5% to 8.6% while Pop increased from 5.6% to 6.8%. That said, in absolute terms, this model is very biased toward better represented genres, despite it's higher overall accuracy.

From this experimentation process, it was found that the Extreme Gradient-Boosted Model was the most effective at classifying music in the FMA, overall. However, like the Random Forest and KNN classifiers, they tended to be very biased toward data that was more represented in the dataset, even with settings like class weighting being enabled. On the other hand, while less predictive overall, the Support Vector Machine was able to be more balanced across different genres, particularly Jazz and Pop where far fewer training examples were provided to the model.

A classification system should have a fair understanding of all genres to be classified, so for the sake of attempting to classify artificially generated music, the Support Vector Machine was used, given it's more balanced learning of all the genres in the dataset, in spite of how unbalanced it was.

@@

Next, machine generated music to test the classifieron was sourced. Given limitations in compute power, having only a 4-core CPU, options were limited in terms of music models that could generate music of a decent quality to be tested. MusicGPT [@musicgpt] is a local application that allows smaller models, such as Meta's MuseGen models [@musegen] to run and generate small clips of music in resource constrained environements. For this setup, the default 'small' model was selected.

@@

These prompts were then fed one by one into the MusicGPT interface, which passed the prompt into the MuseGen model. Each prompt was then set to generate a clip lasting 30 seconds, which mimics the feature extraction setup of @fma. A new chat was created for each genre, to avoid the model confusing the context from the previous prompt with the prompt of the current song being generated. A total of nine .wav files were saved.

@@

As @tbl-comp-results clearly shows, the classifier could not classify any of the machine-generated songs into their respective genres.

It is unclear why the classification performance is so poor. On the one hand, the SVM classifier had an overall performance of 48.9% in the test data from the train/test spit, meaning that it is fairly likely the classifier made a mistake due to not enough training data for certain genres at the very least. The international genre of music is wide-ranging in style and sound and does not necessarily contain repetitive patterns; this could have 'confused' the classifier into learning the patterns of several genres under the same label. So, for example, when an 'Classical' song was passed into the classifier it could have just missed the cutoff by being slightly stylistically dissimilar, causing it to be classified as 'International', the genre of which can serve as a 'catch-all' genre for the classifier model when the features do not fit neatly into the patterns of any of the other genres.

Interestingly, the prompted 'Pop' song was confused for Hip Hop music and the prompted 'Hip Hop' song was confused for Pop music by the classifier, which highlights the difference of how the patterns of these styles were encoded in the classifier and in the music generation model.

Given that the 'small' music generator size was used to generate music of each style, it could be that the music generated did not capture enough the stylistic elements of each genre enough to be classified properly. This is so, because the 'small' generation model size was used and so probably did not have enough training data to effectively capture all the nuances of the music's style. This can be noticed simply by taking a listen to some of the audio samples generated by the Small MusicGPT model and noticing how they do not sound very much like the genre they were instructed to 'quintessentially' describe as the prompts mentioned.

It could also be the the music generation model is not able to capture all the stylistic nuances as a whole, making the structure of the music generation model ineffective at understanding the underlying patterns of music. A comparison between music between a smaller and a larger model would be necessary to see if this is actually the case.

In any case, this setup demonstrates a systematic way of measuring the "human-ness" of a piece of machine-generated music, with the potential of different human-created music datasets, classifiers, music generation models and prompts being interchanged for the ones that are used here in order to measure how well a music generation model has learned a genre or genres of music.

@@

Machine learning methods are also used to recommend music to users within a particular genre or genres. But what makes one song be recommended more than another? While some of it has to do with popularity through the number of clicks, the clicks are ultimately driven through human enjoyment of the particular features of a particular piece of music. Given that features and measurements of popularity can be mined from songs on services like Spotify, it is perfectly reasonable that a savvy cultural analyst could attempt to correlate certain features with high popularity. If these results are then fed into a music generation model, it is conceivable that models could learn to optimize to "please" the recommendation algorithms, creating music that is more popular than any individual human or team of humans could write.

The previous two observations have more general implications for musicology in general. If music can be "optimized" for, like a financial portfolio, or human-like language generation, when does it cease to become music? With the advent of tools that can create music that closely resembles that of humans in a very short amount of time, we are left to ponder what is music? What are aesthetics and beauty in a world largely shaped by algorithms and optimization? Is there a limit to how much machines can mimic human creation before it ceases to be human? These are some of the important questions about auditory aesthetics that need to be considered, experimented on, and addressed as algorithms continue to take over more and more facets of our lives, individual decision making, and agency.

@@

There are several significant weaknesses to this investigation, and so this work  should be mainly considered as scaffolding for future work in this area. First, the small MuseGen Model that was used in this investigation was ineffective at capturing any of the patterns inherent to the genres that it was supposed to describe. Given that it was trained on a very small subset of music data, in order to be able to be computed locally on limited resources, the small model had limited stylistic fidelity to the genres it was supposed to describe and thus the music it generated was unable to be classified properly.

Similarly, the support vector machine classifier had very low overall performance (49%), likely due to the unbalancedness of the training dataset. Given it's poor performance, it is unclear if misclassification was because of the MuseGen modeling failing to generate stylistically faithful music, or whether the classifier was simply ineffective at classification.

Another weakness was that only timbral features were used in the classification process. While @tzanetakis2002 do describe timbral features' efficacy in helping with classification it is important to note that in their paper, rhythmic and harmonic features did improve classification performance in their work. This could have been similarly replicated here and would have likely improved the chances of the classifier being the confouding variable in this setup.

Another important mistake to this setup was the lack of a human listening validation for each of the pieces to act as a baseline. The incorporation of such a step in this setup would have allow this setup to operate on a more experimental basis, and would have lended more weight to the results presented.

Finally, the use of only one generation per genre, combined with a faulty classifier meant that many of the insights drawn from this setup lack statistical power and may not be able to generalize to the general set of songs that could be theoretically generated by the MuseGen model.

These weaknesses present some important next steps in improving this design for more effective inference of the 'human-likeness' of machine generated music. First, larger music generation models, starting with MuseGen, but even incorporating models from other designers, could be used to generate higher fidelity and more stylistically sound audio that could be more easier for a classifier to classify into various genres. This, combined with generating many samples per genre and then each evaluating them would eliminate uncertainty in results and provide more robust insight into the effectiveness of genre pattern learning by these models. Approaching the problem from a different lens, perhaps classification is not the correct way to approach this problem. Perhaps future work in this area could involved comparing the embeddings of classifier models and music generation models and seeing how similar they are. Given that the weights of some models are openly available, perhaps seeing if any of the weights are similar across the two could be made of interest.

Given more compute resources, superior classification models could be trained and tested on. This could involved Convolutional Neural Networks (CNNs), which could be trained on spectrograms of audio files, or even simply the audio files themselves. Larger scale Transformer models could work equally well in this case and could be better at assessing the human-likeness of a machine-generated being of music.

Finally, incorporating the human element into this experiment would be of paramount importance should it be replicated again. Including subjective evalutation of genre by human participants would eliminate the the position of the classifier being the confounding variable and establish a baseline of which to test classification off of.

This setup should be seen as scaffolding for more sophisticated work in attempting to predict how "human-made" a piece of audio is likely to be. With the advent of models such as Suno, and it's accompanying avalanche of AI generated content, the ability to discern between machine-generated and human-made music increases by the day.

>>>>

Recently, AI-powered tools like Suno [@suno] have emerged, targeting consumers interested in "creating" their own music. This raises questions about how machines understand music, especially in the case of generative models such as Suno's. While more research-based proofs of concept like Meta's MuseGen [@musegen] have existed for years, tools like Suno adopt a more consumer-focused approach that inherently makes the algorithms that drive the product even more of a "black box" than before. But how does a word entered into a prompt box translate into the auditory representation humans understand as music? Do different artificial intelligence systems have different conceptualizations of commonly accepted human concepts of genre?

To understand if there is a different between conceptualizations of genre between different machine learning systems, we train several different machine learning algorithms in the task of predicting genre, based on a dataset of the timbral audio features of songs across 10 different genres extracted from the Free Music Archive [@fma]. Then, we ask the 'small' variant of Meta's MuseGen Music Generation Model to create 'quintessential' representations of different genres of music using systematic prompts, and test if the most balanced algorithm can accurately predict the genre based on the extracted audio features.

The most balanced classifier, a Support Vector Machine (SVM) model, was chosen out of four test, but it was unable to guess the genre of any of the machine-generated tracks. It is unclear why this this. It could either be because of the perhaps due to the music generation model's inability to create music that follows patterns of any genre, or it may be because the SVM performs very poorly. In this case, the poor performance of the SVM inhibits deeper insight, but does signal that diffrent machine learning models "learn" the conceptualization of particular music genres and the features that make it up

@@

As @tzanetakis2002, @chen2009, and @bhalke2017 found, timbral metrics allow for sophisticated and parsimonious audio analyses, allowing for efficient classification of genre across various genres of music, so they are made central to the modelling strategy described in @sec-modelling

@@

After understanding timbral features' centrality in genre classification, finding extensive audio feature data to train the classifier was then made of interest. A key requirement of this data is that all music sourced within it must be human generated, and so we use the Free Music Archive [@fma]. The full archive contains 106,574 tracks across 163 subgenres, and was published in 2017. @vaswani2017, the paper describing the general Transformer architecture, and @huang2019, the first paper describing the archictecture of a "music transformer" were published in 2017 and 2018, respectively. Given the close publication time of the dataset and these two publications, it is reasonably unlikely that any artificially-generated music was added to the dataset, since the technology to create sophisticated human-like music did not exist before the dataset was published.

While the FMA is primarily made of audio tracks, which require large amounts of memory to store, Defferard et al. also extracted audio features from each of the tracks using Python's `librosa` module [@librosa], and provided associated metadata for each of the songs, particularly a unique ID, title, artist names, and primary genre of the music.

The data cleaning process began by loading raw audio feature data. From this, the key timbral features of each track were extractedd and consolidated systematically into a single, new dataset. The feature groups that were extracted were the first 16 mean values for the Mel-Frequency Cepstral Coefficients (MFCCs), and the means and variances for the spectral centroid, spectral rolloff, and zero crossing rates for each track. Given that Spectral Flux was not one of the coefficients recorded by Defferard et al, and it is omitted from the feature set as a key predictor.

@@

@fig-genre-bar illustrates the distribution of genre, which is the outcome variable of interest, in the final dataset.

@@

As @fig-genre-bar shows, the dataset exhibits significant class imbalance. Rock dominates with over 14,000 songs, followed by Experimental and Electronic music with approximately 8,000 and 7,000 songs respectively. The remaining 13 genres have substantially fewer observations, with some containing less than 1,000 songs. To mitigate this imbalance and simplify analysis, the dataset was filtered to include only the 10 most frequent genres, as detailed in @tbl-genre-counts.

@@

By reducing the number of classes to 10, the classification task is simplified and more targeted comparisons and analysis can be made between classifying human and machine-generated music. Though the training data is still unbalanced, as seen in @tbl-genre-counts, class-weighting is utilized in the modelling process to account for the overrepresentation of certain genres.

With respect to the predictor variables of the classifier, while it would be possible to visualize each coefficient of the MFCC spectrum, centroids, and other features, each individual coefficient of a song's MFCCs and spectral features are not particularly useful in gleaning special insight on the dataset as a whole. Thus, visualizing the feature set is of adds little to understanding the general aspects of the auditory structure before modelling it and so the features remain unvisualized.

@@

Python's `scikit-learn` [@scikit] and `xgboost` [@xgboost] modules were used to implement several classification models, which were iteratively trained and tested before settling on the one with the best performance. Before training the models, the cleaned dataset was split into 80% training data to 20% testing data. Due to compute limitations and marginal (1-2%) gains in accuracy [@molina2021], cross-validation was not used in the training process, this particular data split allowed for a considerable amount of data to be devoted to training the models, while also leaving some human-made music data to test the performance of the classifiers on. This setup allowed for objective diagnostics of model performance before attempting to classify the artificially-generated music, which can be seen in @sec-model-diag. Wherever possible (all models except KNN), the class-weighting parameter was used to address the imbalanced nature of the genre distribution in the training dataset. This parameter assigns higher penalties to misclassifications of underrepresented classes during model training, helping to mitigate bias toward majority classes. For SVM, Random Forest, and XGBoost classifiers, we applied the 'balanced' weighting option, which automatically adjusts weights inversely proportional to class frequencies. The KNN algorithm does not support class weighting as it makes predictions based solely on distance metrics between data points, rather than through an optimization process.

In order of least to best overall performance, the Support Vector Machine (SVM), K-nearest neighbours, Random Forest, and Extreme Gradient Boosting were evaluated. Due to restrictions in computing power, Deep Learning models were not able to be tested, but were initially considered for this problem. The Support Vector Machine (SVM) model, while the least accurate overall, was the most balanced in prediction across genres. So, it was then used to classify the artificially generated music described in @sec-results.

@@

Support Vector Machines (SVMs) are a widely used supervised learning method for classification tasks, particularly music. SVMs work by finding the best possible boundary, or hyperplane, that separates songs from different genres based on the extracted timbral audio features. When the relationship between genres is too complex for a simple linear boundary, SVMs can apply kernel functions to capture more intricate, non-linear patterns in the data and find higher-dimensional boundaries between the features, effectively classifying them into genres. By focusing mainly on the most informative examples in the dataset, known as support vectors, SVMs achieve strong performance while avoiding overfitting the training data [@islp]. This makes them particularly effective for classifying music, where genre characteristics often overlap and vary in subtle ways.

Both @chen2009 and @bhalke2017 demonstrated particular success in genre classification with SVM, each implementing them to predict genre from audio features to a high degree of accuracy. Bhalke et al. demonstrated nearly 96% accuracy on their music classification task, so it was expected that SVMs would exhibit similar good performance using the FMA data. The setup for our SVM training used the default parameters in `scikit-learn`. That meant our SVM used an RBF kernel with $C=1.0$ for regularization and $gamma='scale'$ to automatically adjust the kernel coefficient based on the input data. The class-weighting parameter was also enabled in order to account for the unbalancedness of the dataset. Results of testing the SVM on the test set are presented in @tbl-svm-results

@@

Surprisingly, this classifier had an accuracy of about 49%, which prompted the investigation of other classification methods for this task described later on. More surprisingly, were the results of testing the SVM on the test data from the FMA. The SVM learned the patterns of Classical music extremely well, despite it having the 2nd least amount of training examples in the dataset, achieving recall of nearly 80% and an F1 score of 0.6. Upon further reflection, this makes sense, as many subgenres of 'Classical' music have particular blends of instruments whose sounds combine together in predictable ways. An example of this within the context of the dataset is that of string quartets, whose timbre of the 2 Violins, Viola, and Cello come together in predictable ways across recordings, interpretations, and composers, which the classifier then learns.

Jazz and Pop genres showed very poor performance, with F1 scores below 20%, despite enabling class-weighting as a parameter. This likely stems from the diverse range of instruments and arrangements in these genres, requiring more training data for effective pattern recognition. The dataset's imbalance is evident: Jazz has only 564 songs and Pop 2,325, compared to Rock's 14,000. Attempts to improve accuracy by selecting the top 10 most important features did not yield significant improvements. Nevertheless, the SVM maintains relatively balanced classification across genres compared to other models tested.

@@

Next, the K-nearest neighbours algorithm (KNN) was tested. @tzanetakis2002 had some modest success using the KNN classifier, with genre classification accuracy above 50% based on only timbral features. KNN classifies a new song based on the genres of its $K$ closest neighbors in the feature space. The number of neighbors, $K$, is a key parameter that balances flexibility and stability: smaller values of K make the model sensitive to local patterns, while larger values smooth out noise but may overlook finer distinctions. KNN does not build an explicit model but instead relies directly on the training data, making it highly adaptable but potentially sensitive to the choice and scaling of features [@islp]. In music genre classification, KNN is particularly useful when songs of the same genre tend to cluster closely together based on their properties. Given that music of a certain genre tends to "sound" similar in terms of timbre, it would make sense that KNN could work well at classifying genre based on how those features cluster together.

The setup for our KNN classifier was such that based on a $K = 25$, basing classification on the 25 nearest neighbours. In repeated trainings, this was found to balance noise control with precision. The results of testing on the test set are presented in @tbl-knn-results

@@

This classifier had an overall accuracy of about 54%, improving it slightly beyond what the SVM could predict overall. Like the SVM, classical music exhibited high performance in the unseen test data, despite low training examples. That said, KNN relies exactly on the training data, there was no class-weighting parameter, so the lesser-represented classes suffered greatly in learning quality. So, from the SVM to KNN, we observed correct classification rates for Pop music go from 16.3% (SVM) to 6.8% (KNN) and Jazz from 14.7% to 3.4%, making this classifier much more biased than the SVM. More represented genres tended to perform even better than in the SVM, with correct Electronic music classification going from 51.9% to 55.3%.

@@

Tp test if classification could be improved further, tree-based classification methods were then attempted. A Random Forest classifier was attempted first.

@@

The Random Forest classifier was trained using class-weighting enabled ('balanced') to address the biased data distribution across genre classes. The model utilized scikit-learn's default hyperparameters, including 100 decision trees (n_estimators=100), considering all features at each split (max_features='sqrt'), no maximum depth limitation (max_depth=None), minimum samples required for a node split of 2 (min_samples_split=2), and minimum samples required at a leaf node of 1 (min_samples_leaf=1). The resulting forest contained trees with a mean of 22,023 nodes and an average depth of 32, indicating relatively complex decision boundaries. Bootstrap samples were used during training (bootstrap=True) with Gini impurity as the splitting criterion (criterion='gini'). The results of testing on the test set are presented in @tbl-rf-results.

@@

Similar to the Random Forest model, the XGBoost classifier was trained with class weighting enabled to address the uneven distribution of genre classes in the dataset. To enhance performance, the model was evaluated using Mean Average Precision (MAP), a metric suited for multiclass classification tasks. Apart from these adjustments, the model relied on XGBoost’s default hyperparameters: 100 estimators, a learning rate of 0.3, a maximum tree depth of 6, and a minimum child weight of 1. The model was trained on the same set of features used for the Random Forest. Ultimately, the XGBoost ensemble was composed of 100 decision trees, collaboratively enhancing prediction accuracy. The results of the model's performance on the test set are shown in Table @tbl-xgb-results.

@@

The XGBoost tree further improved upon the Random Forest in overall accuracy, with a accuracy of 56.4% overall. XGBoost did marginally improve classification rates of the less represented genres of Jazz and Pop. Correct classification of jazz increased from 8.5% to 8.6% while Pop increased from 5.6% to 6.8%. That said, in absolute terms, this model is very biased toward better represented genres, despite it's higher overall accuracy.

From this experimentation process, it was found that the Extreme Gradient-Boosted Model was the most effective at classifying music in the FMA, overall. However, like the Random Forest and KNN classifiers, their classification accuracy tended to be very biased toward data that was more represented in the dataset, even with the class-weighting parameter being enabled. On the other hand, while less predictive overall, the Support Vector Machine was able to be more balanced across different genres, particularly Jazz and Pop where far fewer training examples were provided to the model.

A classification system should have a fair understanding of all genres to be classified, so for the sake of attempting to classify artificially generated music, the Support Vector Machine was used, given the model's balanced learning of all the genres in the dataset.

@@

Next, machine generated music to test the classifieron was sourced. Given limitations in compute power, having only a 4-core CPU, options were limited in terms of music models that could generate music of a decent quality to be tested. MusicGPT [@musicgpt] is a PC-based application that allows smaller models, mainly Meta's MuseGen models [@musegen], to be run and generate small clips of music in resource constrained environments. For this setup, the default 'small' model was selected for the experiment.

@@

These prompts were then fed one by one into the MusicGPT interface, which passed the prompt into the MuseGen model. Each prompt was then set to generate a clip lasting 30 seconds, which mimics the feature extraction setup of @fma. A new chat was created for each genre, to avoid the model confusing the context from the previous prompt with the prompt of the current song being generated and a total of nine .wav files were saved. Notably, despite the classifiers being trained to classify into an 'International' genre, no such song was able to be generated with MusicGPT, since it is too broad a category, genre-wise.

@@

As @tbl-comp-results shows, the classifier could not classify any of the machine-generated songs into their respective genres.

The classification poor results reveal several key insights about both the classifier's performance and the characteristics of the generated music. The SVM classifier misclassified all generated songs, with a striking tendency to categorize multiple tracks as "International". This complete misclassification can be attributed to multiple factors: the classifier's inherent limitations (demonstrated by its modest 48.9% accuracy on human music test data), the "International" genre potentially serving as a catch-all category for songs with ambiguous features due to its broad stylistic range in the training data, and the generated music possibly lacking the genre-specific timbral patterns that the classifier was trained to recognize.

A particularly intriguing observation emerged in the symmetric confusion between Pop and Hip-Hop genres, where the prompted Pop song was classified as Hip-Hop and vice versa. This symmetric misclassification suggests that the small MuseGen model may encode these genres' characteristics differently than they manifest in human-made music, or alternatively, that these genres share similar timbral features that the classifier struggles to distinguish between.

The complete failure to correctly classify any generated songs likely stems from the limitations of using MuseGen's "small" model variant. This model, because it was designed for resource-constrained environments, may lack the capacity to capture and reproduce finer genre-specific timbral patterns effectively. This limitation is apparent even to human listeners, as the generated samples often lack clear genre-specific characteristics despite being prompted to be "quintessential" examples.

From these results, there are two possible explanations: either the classifier's poor performance (48.9% accuracy) makes it unreliable, or the generated music fundamentally differs from human-made music in its timbral features.

This setup demonstrates a workflow for assessing AI music generation, but the SVM's limitations make it difficult to determine if misclassification stems from the classifier or fundamental differences in the generated music. Future work should use higher-performing classifiers, test larger music generation models, and incorporate human validation.

Despite limitations, this experiment provides a systematic approach for measuring how well a music generation model learns genre-specific characteristics. The methodology could be improved through better, more balanced datasets,  classifiers with stronger performance (above 80% at least), larger models, or more sophisticated prompting.

@@

Machine learning methods are also used to recommend music to users within a particular genre or genres. But what makes one song be recommended more than another? While some of it has to do with popularity through the number of clicks, the clicks are ultimately driven through human enjoyment of the particular features of a piece of music. Given that features and measurements of popularity can be mined from songs on services like Spotify, it is perfectly reasonable that a savvy cultural analyst could attempt to correlate certain features with high popularity. If these results are then fed into a music generation model, it is conceivable that models could learn to optimize to "please" the recommendation algorithms, creating music that is more popular than any individual human or team of humans could write.

The previous two observations have more general implications for musicology in general. If music can be "optimized" for, like a financial portfolio or human-like language generation, when does it cease to become music? As tools emerge that can generate music nearly indistinguishable from that created by humans - and in a fraction of the time - we are prompted to reconsider the question: what is music? What are aesthetics and beauty in a world largely shaped by algorithms and optimization? Is there a limit to how much machines can mimic human creation before it ceases to be human altogether? These are some of the important questions about auditory aesthetics that need to be considered, experimented on, and addressed by humans, as algorithms continue to take over more and more facets of our lives, individual decision making, and agency.

@@

This investigation has several significant weaknesses that should be considered when interpreting the results, suggesting that this work primarily serves as scaffolding for future research in this domain.

The small MuseGen model employed in this study demonstrated limited capability in capturing genre-specific patterns. Trained on a restricted subset of music data to accommodate local computational constraints, the model produced outputs with poor stylistic fidelity to the intended genres. Consequently, these generated pieces lacked the distinctive characteristics necessary for accurate genre classification.

The Support Vector Machine (SVM) classifier exhibited inadequate performance (49% accuracy), likely attributable to the imbalanced training dataset. This poor classification rate introduces ambiguity regarding whether misclassifications stemmed from the MuseGen model's failure to generate stylistically faithful music or from the classifier's inherent limitations in distinguishing musical genres.

A methodological limitation was the exclusive reliance on timbral features for classification. While @tzanetakis2002 demonstrated the efficacy of timbral features in genre classification, their research also highlighted that incorporating rhythmic and harmonic features significantly improved classification performance. The absence of these additional feature types in our approach, along with the omission of important timbral aspects like Spectral Flux (due to limitations in the FMA dataset), likely constrained classification accuracy.

The absence of human validation represents another critical shortcoming in our experimental design. Without human listeners evaluating the genre categorization of the generated pieces, we lacked a baseline against which to measure both the classifier's performance and the stylistic authenticity of the AI-generated music. Such validation would have strengthened the experimental foundation and added weight to our findings.

Finally, the limited scope of generating only one sample per genre, combined with the aforementioned classifier limitations, undermines the statistical power of our results. This approach precludes generalization to the broader set of potential outputs from the MuseGen model, restricting the validity and applicability of our conclusions regarding machine-generated music.

Building upon the limitations identified, several concrete directions for future research emerge that could substantially enhance our understanding of machine-generated music classification and evaluation.

Future studies should employ larger, more sophisticated music generation models. Specifically, experiments could compare outputs from the full-scale MuseGen model against those from alternative frameworks such as Suno, MusicLM, and MusicGen. A comparative analysis of these models' outputs—using the same genre prompts—would provide insights into which architectures produce the most stylistically faithful music across different genres. This approach would help identify whether certain models excel at specific genres while struggling with others.

A more robust experimental design should incorporate multiple generated samples per genre (e.g., 25-50 samples) to enable statistical analysis of classification performance. Each sample could be analyzed individually and collectively to determine if certain genres consistently prove more challenging for both generation and classification. This expanded dataset would allow for meaningful statistical testing and more defensible conclusions about model performance.

Future work should explore alternative classification approaches beyond traditional machine learning algorithms. Specifically, implementing Convolutional Neural Networks (CNNs) trained directly on audio spectrograms could capitalize on spatial patterns within the frequency domain, which were shown by @dong2018 to have at least a human-level (70%) accuracy, if not higher with current developments in model tuning. Additionally, transformer-based classifiers could potentially capture longer-term dependencies in the musical structure. A direct comparison between these approaches and traditional methods (SVM, Random Forest, KNN) using standardized evaluation metrics would quantify the performance improvements offered by more advanced techniques.

The feature extraction process should be expanded to include rhythmic and harmonic features alongside timbral characteristics. Concrete implementations could include beat histogram analysis for rhythmic features and chord progression detection for harmonic content. Following the methodology @tzanetakis2002 more comprehensively by implementing their full feature set could likely improve classification performance and provide a more nuanced understanding of genre distinctions in AI-generated music.

A critical enhancement would be through incorporating human evaluation through structured listening tests. A formal experiment could involve music experts rating generated samples on genre adherence using Likert scales and providing qualitative feedback. This human-centered evaluation could serve as a gold standard against which to measure classifier performance. Additionally, A/B testing that asks participants to distinguish between human-composed and AI-generated music within specific genres would provide insights into which genres machines replicate most convincingly.

Finally, future research could explore embedding similarity between classifier models and music generation models. This novel approach would involve extracting and comparing the latent representations from both model types when processing the same musical inputs. If publicly available, comparing the weight distributions across model layers could reveal whether certain musical features are encoded similarly in both classification and generation systems. This analysis might uncover fundamental commonalities in how machines represent musical concepts, regardless of whether the task is generative or discriminative.

Collectively, these proposed directions address the limitations of the current study while advancing our understanding of how machines conceptualize and reproduce musical genres. With the rapid proliferation of AI-generated music through platforms like Suno, developing reliable methods to analyze and evaluate such content becomes increasingly important for both academic research and practical applications in the music industry.
